Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-10-07 paper.Rnw
updates to paper after meeting (just need to proof read sections 4 and 5 one more time)

The explanation scatterplot is built by plotting the LIME simulated data for the top two features identified by the explanation in a scatterplot and coloring these points by the  predictions from the complex model. The point size  represents the weight assigned by LIME. In order to show the LIME results for the observation of interest, lines are drawn on top of the points. These lines  represent the boundaries of the indicator variables used to fit the explainer model. The line color  denotes whether LIME indicates that a feature supports or contradicts a class prediction. Appendix \ref{exp-scatter_plus} addresses the explanation scatterplot formats in other LIME simulation scenarios.

@@

%XXX minor change to emphasize that this is not a completely new model, just an expanded training set - otherwise you open yourself up to criticism of not using a validated model. The bulletxtrctr model was used in an external validation in \citet{vanderplas:2020b}.

\kgc{Your comment about being clear that we are not training a whole new model made lots of sense. I made a few more changes. Do they look okay?} \kge{A random forest model fit to an expanded dataset of \Sexpr{format(dim(bullet_train)[1], big.mark = ",")} land-to-land comparisons from two sets of bullets in the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009} was validated in \citet{vanderplas:2020b}. Here, we train a random forest model to mimic the model validated in \citet{vanderplas:2020b} using the same data, model structure, and features (the nine similarity measures developed in \citet{hare:2017})}. %Here, we train a random forest model using the same \kge{data, features (the nine similarity measures developed in \citet{hare:2017}), and model structure as a random forest validated in \citet{vanderplas:2020b}} \kge{\sout{nine comparison features as \citet{hare:2017}}} to predict whether two lands come from the same source (i.e.\ are a match) in an expanded set of \Sexpr{format(dim(bullet_train)[1], big.mark = ",")} land-to-land comparisons from two sets of bullets in the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}. 
We apply the trained random forest model to 6 bullets from another set of the Hamby study with \Sexpr{dim(bullet_test)[1]} rows of land comparisons. See the supporting information for further descriptions of the data. \kge{The newly trained random forest has an out-of-bag accuracy of \Sexpr{bullet_rf_oob_acc} and out-of-bag false positive and false negative rates of \Sexpr{bullet_rf_oob_fp} and \Sexpr{bullet_rf_oob_fn}, respectively. On the test data, the random forest has an accuracy of \Sexpr{bullet_rf_test_acc} and false positive and false negative rates of \Sexpr{bullet_rf_test_fp} and \Sexpr{bullet_rf_test_fn}, respectively.}

We first consider a global visualization of the relationship between the random forest predictions and the model features with a parallel coordinate plot of the training data (top facet) and testing data (bottom facets)  (\autoref{fig:figure-10}).  In the training data we observe a clear difference in feature values based on whether the corresponding random forest scores are close to 1 or 0. High values of rough correlation, cross correlation function, and number of matches are indicative of random forest scores close to 1; similarly, low values of mismatches, distance, and non-consecutively matching striae are also associated with random forest scores close to 1. Observations in the test data that are classified incorrectly by the random forest tend to have feature values similar to observations in the training data with similar random forest scores. For example, the observations in the test set known to be a match but have a random forest score below 0.5 have relatively low values of rough correlation, cross correlation function, and matches and relatively large values of mismatches, distance, and non-consecutively matching striae compared to observations with a random forest probability close to 1.

@@

\kge{We \sout{First,}} focus on the explanation for case NM with 4-quantile-bins. The explanation plot from \emph{lime} shows that \kge{the} observed values \kge{of NM of \sout{with}} \kge{mismatches less than or equal to 7.17, rough correlation (rough\_cor) greater than 0.05, and D less than or equal to 0.002 support a random forest prediction in favor of a match. The scatter plots show that many of the simulated values in the rectangular regions with mismatches less than 7.17, rough correlation greater than 0.05, and distance less than or equal to 0.002 have random forest predictions greater than 0.5.} %rough correlation (rough\_cor) greater than 0.05 and mismatches value less than or equal to 7.17 support a random forest prediction in favor of a match, and the scatter plots show that many of the simulated values with rough correlation greater than 0.05 and mismatches less than 7.17 have random forest predictions greater than 0.5. Additionally, the LIME explanation indicates that a value of cross correlation function (ccf) greater than 0.28 and less than or equal to 0.35 supports a non-match, and the scatter plots shows that the simulated values within this region have mostly random forest predictions less than 0.5. 
However, the random forest assigns a probability of 0.21 to case NM, which is not explained by the LIME explanation \kge{that provides all explanations in support of a match}. Instead, consider that a better explanation based on the scatterplots would be that the random forest assigned a prediction probably \kge{below 0.5 \sout{close to 0}} since \kge{NM has observed values of mismatches less than 5, rough correlation greater than 0.25, and distance greater than 0.001.}%the observation has a rough correlation less than 0.5, mismatches greater than 4, and cross correlation function 0.6.

The other explanation scatterplots can be interpreted in a similar manner. \kge{\sout{Overall, t} T}he \kge{three plots depict situations where the explanations are faithful} %plots identify that the 4-quantile-bin explanations are not faithful to the random forest model predictions. The 4-equal-bin explanations are more faithful 
to the random forest, but the regions captured by the \kge{intersections of the} bins do not align well with the regions containing similar probabilities produced by the random forest.

@@

Even though an explanation will be interpretable as long as the complexity of the explainer model is appropriately chosen, a lack of understanding of the process used to create the explanation could lead to an incorrect interpretation of the explanation. For example, the visualization of a LIME explanation available from the \emph{lime} R package \citep{pedersen:2020} (shown in Figures \ref{fig:figure-04} and \ref{fig:figure-13}) is a major simplification of the explainer model, which could lead to under-interpreted or misinterpreted LIME explanation. Supplementing \citet{pedersen:2020}'s compact visualization of the explanation with an explanation scatterplot that shows a more detailed visualization of the explainer model (such as Figures~\ref{fig:figure-04} and \ref{fig:figure-13}) promotes a more complete understanding of the explanation.
Even with an explainer model that is interpreted correctly, the interpretation is worthless if the explainer model is not \textbf{faithful} to the complex model. This claim can be assessed using the diagnostic plots suggested in this paper. Many of the visualizations in this paper highlight problems with the faithfulness of the explainer models. The explanation scatterplots allow for a comparison of the explainer model to the complex model. The examples in this paper show cases where the explainer model bins do not accurately capture the regions with similar random forest probabilities that contain the prediction of interest and oversimplify the model (Figures \ref{fig:figure-04} and \ref{fig:figure-13}). Using fewer bins would clearly not help improve the faithfulness of the explainer model in these examples, and while an increase in the number of bins would lead to a finer resolution of the random forest classification boundaries, interpretability of the explainer model would quickly be lost. Perhaps this could be improved by allowing the bin creation to account for the relationships between the features and response variable or a different number of bins for each feature.

>>>>

The explanation scatterplot applies the concept of plotting the model in the data space discussed in \citet{wickham:2015}. The plot is built by plotting the LIME simulated data for the top two features identified by the explanation in a scatterplot and coloring these points by the predictions from the complex model. The point size represents the weight assigned by LIME. In order to show the LIME results for the observation of interest, lines are drawn on top of the points. These lines  represent the boundaries of the indicator variables used to fit the explainer model. The line color  denotes whether LIME indicates that a feature supports or contradicts a class prediction. Appendix \ref{exp-scatter_plus} addresses the explanation scatterplot formats in other LIME simulation scenarios.

@@

A random forest model fit to an expanded dataset of \Sexpr{format(dim(bullet_train)[1], big.mark = ",")} land-to-land comparisons from two sets of bullets in the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009} was validated in \citet{vanderplas:2020}. Here, we train a random forest model to mimic the model validated in \citet{vanderplas:2020} using the same data, model structure, and features (the nine similarity measures developed in \citet{hare:2017}). We apply the trained random forest model to 6 bullets from another set of the Hamby study with \Sexpr{dim(bullet_test)[1]} rows of land comparisons. See the supporting information for further descriptions of the data.

The newly trained random forest has an out-of-bag accuracy of \Sexpr{bullet_rf_oob_acc} and out-of-bag false positive and false negative rates of \Sexpr{bullet_rf_oob_fp} and \Sexpr{bullet_rf_oob_fn}, respectively. On the test data, the random forest performance decreases with an accuracy of \Sexpr{bullet_rf_test_acc} and false positive and false negative rates of \Sexpr{bullet_rf_test_fp} and \Sexpr{bullet_rf_test_fn}, respectively. This is an example where explanations of the model predictions could provide insight to the cause of the decrease in model performance on the test data.

We consider a global visualization of the relationship between the random forest predictions and the model features with a parallel coordinate plot of the training data (top facet) and testing data (bottom facets)  (\autoref{fig:figure-10}). In the training data we observe a clear difference in feature values based on whether the corresponding random forest scores are close to 1 or 0. High values of rough correlation, cross correlation function, and number of matches are indicative of random forest scores close to 1; similarly, low values of mismatches, distance, and non-consecutively matching striae are also associated with random forest scores close to 1. Observations in the test data that are classified incorrectly by the random forest tend to have feature values similar to observations in the training data with similar random forest scores. For example, the observations in the test set known to be a match but have a random forest score below 0.5 have relatively low values of rough correlation, cross correlation function, and matches and relatively large values of mismatches, distance, and non-consecutively matching striae compared to observations with a random forest probability close to 1.

@@

We focus on the explanation for case NM with 4-quantile-bins. The explanation plot from \emph{lime} shows that the observed values of NM of mismatches less than or equal to 7.17, rough correlation (rough\_cor) greater than 0.05, and D less than or equal to 0.002 support a random forest prediction in favor of a match. The scatter plots show that many of the simulated values in the rectangular regions with mismatches less than 7.17, rough correlation greater than 0.05, and distance less than or equal to 0.002 have random forest predictions greater than 0.5. However, the random forest assigns a probability of 0.21 to case NM, which is not explained by the LIME explanation that provides all explanations in support of a match. Instead, consider that a better explanation based on the scatterplots would be that the random forest assigned a prediction probably below 0.5 since NM has observed values of mismatches less than 5, rough correlation greater than 0.25, and distance greater than 0.001.

The other explanation scatterplots can be interpreted in a similar manner. The three plots depict situations where the explanations are faithful to the random forest, but the regions captured by the intersections of the bins do not align well with the regions containing similar probabilities produced by the random forest.

@@

Even though an explanation will be interpretable as long as the complexity of the explainer model is appropriately chosen, a lack of understanding of the process used to create the explanation could lead to an incorrect interpretation of the explanation. For example, the visualization of a LIME explanation available from the \emph{lime} R package \citep{pedersen:2020} (shown in Figures \ref{fig:figure-04} and \ref{fig:figure-13}) is a major simplification of the explainer model, which could lead to under-interpreted or misinterpreted LIME explanation. Supplementing \citet{pedersen:2020}'s compact visualization of the explanation with an explanation scatterplot that shows a more detailed visualization of the explainer model (such as Figures~\ref{fig:figure-04} and \ref{fig:figure-13}) promotes a more complete understanding of the explanation.

Even with an explainer model that is interpreted correctly, the interpretation is worthless if the explainer model is not \textbf{faithful} to the complex model. This claim can be assessed using the diagnostic plots suggested in this paper. Many of the visualizations in this paper highlight problems with the faithfulness of the explainer models. The explanation scatterplots allow for a comparison of the explainer model to the complex model. The examples in this paper show cases where the explainer model bins do not accurately capture the regions with similar random forest probabilities that contain the prediction of interest and oversimplify the model (Figures \ref{fig:figure-04} and \ref{fig:figure-13}). Using fewer bins would clearly not help improve the faithfulness of the explainer model in these examples, and while an increase in the number of bins would lead to a finer resolution of the random forest classification boundaries, interpretability of the explainer model would quickly be lost. Perhaps this could be improved by allowing the bin creation to account for the relationships between the features and response variable or a different number of bins for each feature.

### 2020-10-07 paper.tex
updates to paper after meeting (just need to proof read sections 4 and 5 one more time)

\hh{XXX minor change to emphasize that this is not a completly new model, just an expanded training set - otherwise you open yourself up to criticism of not using a validated model. The bulletxtrctr model was used in an external validation in \citet{validation}. }
\hh{Here, we train a } random forest model using the same nine comparison features as \citet{hare:2017} to predict whether  \hh{two lands come from the same source} (i.e.\ are a match) in an expanded set of 83,028 land-to-land comparisons from \hh{two sets of} bullets in the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}. We apply the trained random forest model to 6 bullets from \hh{another set of} the Hamby study with 364 rows of land comparisons. See the supporting information for further descriptions of the data.

\hh{XXX mention overall goodness of fit or oob error rates?}
We first consider a global visualization of the relationship between the random forest predictions and the model features with a parallel coordinate plot of the training data (top facet) and testing data (bottom facets)  (\autoref{fig:figure-10}).  In the training data we observe a clear difference in feature values based on whether the corresponding random forest scores are close to 1 or 0. High values of rough correlation, cross correlation function, and number of matches are indicative of random forest scores close to 1; similarly, low values of mismatches, distance, and non-consecutively matching striae are also associated with random forest scores close to 1.
Observations in the test data that are classified incorrectly by the random forest tend to have feature values similar to observations in the training data with similar random forest scores. For example, the observations in the test set known to be a match but have a random forest score below 0.5 have relatively low values of rough correlation, cross correlation function, and matches and relatively large values of mismatches, distance, and non-consecutively matching striae compared to observations with a random forest probability close to 1.

@@

First, focus on the explanation for case NM with 4-quantile-bins. The explanation plot from \emph{lime} shows that observed values with rough correlation (rough\_cor) greater than 0.05 and mismatches value less than or equal to 7.17 support a random forest prediction in favor of a match, and the scatter plots show that many of the simulated values with rough correlation greater than 0.05 and mismatches less than 7.17 have random forest predictions greater than 0.5. Additionally, the LIME explanation indicates that a value of cross correlation function (ccf) greater than 0.28 and less than or equal to 0.35 supports a non-match, and the scatter plots shows that the simulated values within this region have mostly random forest predictions less than 0.5.  However, the random forest assigns a probability of 0.19 to case NM, which is not explained by the LIME explanation. Instead, consider that a better explanation based on the scatterplots would be that the random forest assigned a prediction probably close to 0 since the observation has a rough correlation less than 0.5, mismatches greater than 4, and cross correlation function 0.6.

The other explanation scatterplots can be interpreted in a similar manner. Overall, the plots identify that the 4-quantile-bin explanations are not faithful to the random forest model predictions. The 4-equal-bin explanations are more faithful to the random forest, but the regions captured by the bins do not align well with the regions containing  similar probabilities produced by the random forest.

>>>>

A random forest model fit to an expanded dataset of 83,028 land-to-land comparisons from two sets of bullets in the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009} was validated in \citet{vanderplas:2020}. Here, we train a random forest model to mimic the model validated in \citet{vanderplas:2020} using the same data, model structure, and features (the nine similarity measures developed in \citet{hare:2017}). We apply the trained random forest model to 6 bullets from another set of the Hamby study with 364 rows of land comparisons. See the supporting information for further descriptions of the data.

The newly trained random forest has an out-of-bag accuracy of 1 and out-of-bag false positive and false negative rates of 0.3 and \ensuremath{2\times 10^{-4}}, respectively. On the test data, the random forest performance decreases with an accuracy of 0.85 and false positive and false negative rates of 0.55 and 0.011, respectively. This is an example where explanations of the model predictions could provide insight to the cause of the decrease in model performance on the test data.

We consider a global visualization of the relationship between the random forest predictions and the model features with a parallel coordinate plot of the training data (top facet) and testing data (bottom facets)  (\autoref{fig:figure-10}). In the training data we observe a clear difference in feature values based on whether the corresponding random forest scores are close to 1 or 0. High values of rough correlation, cross correlation function, and number of matches are indicative of random forest scores close to 1; similarly, low values of mismatches, distance, and non-consecutively matching striae are also associated with random forest scores close to 1. Observations in the test data that are classified incorrectly by the random forest tend to have feature values similar to observations in the training data with similar random forest scores. For example, the observations in the test set known to be a match but have a random forest score below 0.5 have relatively low values of rough correlation, cross correlation function, and matches and relatively large values of mismatches, distance, and non-consecutively matching striae compared to observations with a random forest probability close to 1.

@@

We focus on the explanation for case NM with 4-quantile-bins. The explanation plot from \emph{lime} shows that the observed values of NM of mismatches less than or equal to 7.17, rough correlation (rough\_cor) greater than 0.05, and D less than or equal to 0.002 support a random forest prediction in favor of a match. The scatter plots show that many of the simulated values in the rectangular regions with mismatches less than 7.17, rough correlation greater than 0.05, and distance less than or equal to 0.002 have random forest predictions greater than 0.5. However, the random forest assigns a probability of 0.21 to case NM, which is not explained by the LIME explanation that provides all explanations in support of a match. Instead, consider that a better explanation based on the scatterplots would be that the random forest assigned a prediction probably below 0.5 since NM has observed values of mismatches less than 5, rough correlation greater than 0.25, and distance greater than 0.001.

The other explanation scatterplots can be interpreted in a similar manner. The three plots depict situations where the explanations are faithful to the random forest, but the regions captured by the intersections of the bins do not align well with the regions containing similar probabilities produced by the random forest.
